\documentclass[a4paper,11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsthm,amsopn}
\usepackage{mathrsfs}
\usepackage{graphicx}
%\usepackage{tikz}
%\usepackage{array}
%\usepackage[top=1cm,bottom=1cm]{geometry}
%\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=red,
}

\newtheoremstyle{break}%
{}{}%
{\itshape}{}%
{\bfseries}{}%  % Note that final punctuation is omitted.
{\newline}{}

\newtheoremstyle{sc}%
{}{}%
{}{}%
{\scshape}{}%  % Note that final punctuation is omitted.
{\newline}{}

\theoremstyle{break}
\newtheorem{thm}{Theorem}[section]
\newtheorem{lm}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{cor}[thm]{Corollary}

\theoremstyle{sc}
\newtheorem{exo}{Exercise}

\theoremstyle{definition}
\newtheorem{defi}[thm]{Definition}
\newtheorem{ex}[thm]{Example}

\theoremstyle{remark}
\newtheorem{rem}[thm]{Remark}

\DeclareMathOperator{\Ker}{Ker}
\DeclareMathOperator{\Id}{Id}
\DeclareMathOperator{\Img}{Im}
\DeclareMathOperator{\Card}{Card}
\DeclareMathOperator{\Vect}{Vect}
\DeclareMathOperator{\Tr}{Tr}


% Nouvelles commandes
\newcommand{\ps}[2]{\left\langle#1,#2\right\rangle}
\newcommand{\ent}[2]{[\![#1,#2]\!]}
\newcommand{\diff}{\mathop{}\!\mathrm{d}}
\newcommand{\ie}{\emph{i.e. }}
\newcommand{\q}{\textcolor{violet}{\textbf{ (???) }}} % when I do not really
% know what is going on or when I still need to ckeck

% opening
\title{}
\author{The powers-of-2 descent\\\textcolor{violet}{\textbf{Work in progress}}}


\begin{document}

\maketitle

%\begin{abstract}

%\end{abstract}

%\tableofcontents

%\clearpage

\section{Introduction}

The discrete logarithm problem has seen dramatic improvements these last
years in the small characteristic case~\cite{Joux13, BGJT13, GKZ14},
culminating with two \emph{quasi-polynomial} algorithms in 2014. ``On the powers
of $2$'' is the paper~\cite{GKZ14} introducing one of those two algorithms and this document is
meant to give some explanations about it. Before going into details about this
algorithm, we recall a few facts about discrete logarithm. A more complete
panorama can be found in suveys such as~\cite{JP16, GKZ16}.

\section{The Discrete Logarithm Problem (DLP)}
Let $\mathcal G=\left\langle g\right\rangle$ be a cyclic group generated by an element
$g$, and denote by $N=|\mathcal G|$ its cardinal. We have the isomorphism:
\[
 \begin{array}{cccc}
   exp_g: & \mathbb{Z}/N\mathbb{Z} & \to & \mathcal G \\
   & n & \mapsto & g^n,
 \end{array}
\]
and we denote by $\log_g=\exp_g^{-1}$ the inverse isomorphism. In practice, the
\emph{square and multiply} algorithm allows us to compute $g^n=\exp_g(n)$
efficiently, \ie in polynomial time. But, given $y = g^k$, the computation of $k
= \log_g(y)$ is not as easy. This kind of function $f$, where $f$ is easy to
compute but $f^{-1}$ is hard to compute, are called \emph{one-way} functions. They are
typically used in cryptology to make the encryption fast and the deciphering
slow. The DLP first appeared in the article of Diffie and
Hellman in 1976, ``New Directions in Cryptography''~\cite{DH76}. It was supposed to be a very hard problem,
\ie only an exponential time algorithm was known at that time, and all the
security of the protocol invented in this article relied on the hardness of the
DLP. This article attracted a lot of attention, and so did the DLP, which has
been widely studied since then. Originally, the group involved in the Diffie-Hellman
protocol was $(\mathbb{Z}/N\mathbb{Z})^\times$, but there are other interesting
groups, such as the points of an elliptic curve or the invertible element of a
finite field. Over the years, two types of algorithms emerged:
\begin{itemize}
  \item the \emph{generic} algorithms, that can be used for any group, with an
    exponential complexity of type $O(\sqrt N)$, where $N=|\mathcal G|$ is the cardinal
    of the group;
  \item the \emph{index calculus} algorithms, that are built on the structure of
    the group considered, and that have very different complexities, depending
    on the kind of the studied group.
\end{itemize}

The recent algorithms that achieve quasi-polynomial complexity are index
calculus algorithms, so we introduce these algorithms a bit more.

\section{The index calculus method}

Here again, we consider a group $\mathcal G=\left\langle g \right\rangle$, an element
$g^k = y\in \mathcal G$, and we want to compute $k$. The index calculus algorithms always
follow the same pattern:
\begin{enumerate}
  \item[0.] we choose $\mathcal F\subset \mathcal G$ such that $\left\langle
    \mathcal F \right\rangle = \mathcal G$
    (we often have $g\in \mathcal F$), this subset is called the \emph{factor base};
  \item we generate multiplicative relations between the elements of $\mathcal F$, \ie we
    find $f_1, \dots, f_n \in \mathcal F$ and $e_1, \dots, e_n\in \mathbb{Z}$ such that
    $\prod_i f_i^{e_i} = 1$, that is equivalent to the linear equation $\sum_i
    e_i\log_g(f_i) = 0$;
  \item we solve the linear system with unknowns $\log_g(f_i)$ arising from step
    1 to obtain $\log(f)$ for all $f\in \mathcal F$;
  \item we find a multiplicative relation between $y$ and elements of $\mathcal F$, or
    equivalently, we express $\log_g y = k$ as a linear combination of the
  $\log_g(f_i)$.
\end{enumerate}
\textcolor{red}{Example : Hellman-Reyneri algorithm, like in the slides for
  F.Paulin ?}
  \newline
  \newline
  An \textcolor{red}{(other ?)} example is given by the algorithm of Granger,
  Kleinjung and Zumbrägel in~\cite{GKZ14}, that we call powers-of-2 algorithm here.
\section{The powers-of-2 algorithm}
\subsection{Setup of the algorithm}

Let $\mathcal G = (\mathbb{F}_{q^n})^\times$ the group of invertible elements of a finite
field of small characteristic $\mathbb{F}_{q^n}$, where $q=p^l$ is a prime
power, and we represent elements of $\mathbb{F}_{q^n}$ by polynomials over
$\mathbb{F}_q$ of degree less than $n$. The setup in which we present the algorithm is slightly different from
the original, thanks to the work of Joux and Pierrot in \cite{JP14}. Indeed,
they remarked that working with polynomials of degree up to $d$ over $\mathbb{F}_q$ is
essentially equivalent as working with linear polynomials over $\mathbb{F}_{q^d}$. Thus,
instead of working with a factor base composed of linear polynomials over
$\mathbb{F}_{q^d}$ (with $d$ usually between $2$ and $4$), we work with a
factor base composed of irreducible polynomials up to degree $4$. Joux and Pierrot
precisely give a way to compute such a factor base in~\cite{JP14}, that is based
on previous work~\cite{Joux13, BGJT13} and has a complexity of $O(q^6)$. Since
we work in the case where $q$ is polynomial in the bitsize of the field
$\log(q^n)$, this precomputation time is polynomial.

Let us come back to the elements of $\mathbb{F}_{q^n}$, we said that they were
represented by polymomials of degree at most $n-1$ over $\mathbb{F}_q$. More
precisely, we consider $\mathbb{F}_{q^n}\cong \mathbb{F}_{q}[X]/(P)$ where $P\in
\mathbb{F}_{q}[X]$ is an irreducible polynomial of degree $n$ over
$\mathbb{F}_{q}$ dividing $h_1X^q-h_0$, and $h_0, h_1\in \mathbb{F}_q[X]$ are
polynomials of degree at most $2$ over $\mathbb{F}_q$. Different options are
available to choose $h_0$ and $h_1$, and the respective results are also
discussed in~\cite{JP14}. The existence of such parameters $h_0$ and $h_1$ is
critical for the algorithm, and the fact that every finite field admits such a
structure is \emph{heuristic}, and remains one of the major open problems in the
area.

We now assume that the factor base $\mathcal F$ is composed of all the
elements of $\mathcal G$ represented by an irreducible polynomial of degree at
most $4$. We also suppose that we know the logarithm of each element in
$\mathcal F$. In other words, we assume that the steps $1$ and $2$ of the index
calculus method have already been performed. The only remaining step is $3$,
where we express the logarithm of an element in $\mathcal G$ as the linear
combination of the logarithms of elements in $\mathcal F$. In order to perform
that step, we express the logarithm of element that we want to study, denoted by
$Q\in\mathcal G$, as a linear combination of ``simpler'' $q+2$ elements $Q_j$,
\ie elements represented by polynomials of degree two times smaller. We
still note $Q$ for the polynomial representing $Q\in
\mathcal G$, and so have $\deg Q_i\leq \lceil \frac{1}{2}\deg Q\rceil$. Doing
the same process reccursively, we are able to express $\log Q$ as a linear
combination of elements of degree at most $4$, \ie elements in $\mathcal F$.
This whole procedure is called the \emph{descent}, and we explain it in the
following.

\subsection{On-the-fly degree $2$ elimination}

The crucial part of the descent is called the called the \emph{on-the-fly}
elimination. It allows one to express a degree $2$ polynomial
$Q\in\mathbb{F}_{q^m}[X]$ over $\mathbb{F}_{q^m}$ as a product of degree $1$ 
polynomials over $\mathbb{F}_{q^m}$. The name ``on the fly'' is due to the fact
that no relation gathering or linear algebra is performed during this
elimination. At the core of this elimination we also use the fact that the
polynomial $\mathcal P_B = X^{q+1}-BX+B$ splits completely over $\mathbb{F}_{q^m}$ (\ie is a
product of polynomial of degree $1$ over $\mathbb{F}_{q^m}$) very often. Let
$\mathcal B$ be the set of elements $B\in \mathbb{F}_{q^m}$ such that $\mathcal
P_B$ splits completely. In~\cite{Bluher04}, the roots and the splitting field of
$\mathcal P_B$ are studied and it is shown that the cardinality of $\mathcal B$
is approximately $q^{m-3}$ ($(q^{m-1}-1)/(q^2-1)$ when $m$ is odd,
$(q^{m-1}-q)/(q^2-1)$ when $m$ is even). Thus, if $B\in \mathcal B$,
$a,b,c\in\mathbb{F}_{q^m}$ such that $c\neq ab$, $a^q\neq b$, we see by the change of variable
$X\to\frac{c-ab}{a^q-b}X-a$ in $\mathcal P_B$ that the polynomial
$X^{q+1}+aX^q+bX+c$ also splits completely \q. Note that using~\cite{Bluher04,
HK10, GKZ14}, we also know that $\mathcal B$ is the image of
$\mathbb{F}_{q^k}\setminus\mathbb{F}_{q^2}$ under the map 
\[
  u\mapsto \frac{(u-u^{q^2})^{q+1}}{(u-u^q)^{q^2+1}}.\q
\]
\subsection{The descent}

\section{A few comments}

In practical works, summarised in Gora Adj PhD thesis~\cite{Adj16}, we can see that this algorithm in not really used...

\clearpage
\bibliographystyle{plain}
\bibliography{dlog}
\end{document}
